---
title: "Be Ultrametric, damnit!"
author: "Christoph Liedtke"
date: "14/08/2019"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Be ultrametric, damnit!

I ran into this problem vairous times when downloading a time-calibrated phylogeny off a repository and expecting it to be ultrametric. *ape's* `is.ultrametric()` kicks back a `FALSE` and for some downstream analyses, functions won't accept it. This drove me crazy because the trees look pretty damn ultrametric. It took me a while to figure out that either just one or two terminal branches are protruding a tiny, tiny bit, or that the distance from root to tip vaires between tips by the tiniest decimal.  

My solution to this problem was to just clip/truncate or elongate the terminal branches so that the root-to-tip distance is the same for all tips.  

This works pretty well for me and only needs the *ape* and *adephylo* package.  

Here is an example:

#### Make a random ultrametric tree, and then break it (for demonstrative purposes)
```{r }
library(ape)
library(adephylo)


## make a random ultrametric tree
set.seed(12345)
tree<-rcoal(50)
par(mfrow=c(1,2))
plot(tree, show.tip.label = F, main="random ultrametric tree")

## lets change some branch lengths by a tiny amount that you can't even see:
tree$edge.length[50:53]<-tree$edge.length[50:53]+0.0095
tree$edge.length[60:62]<-tree$edge.length[60:62]-0.0002
plot(tree, show.tip.label = F, main="non-ultrametric tree")
is.ultrametric(tree)
```


#### Make it ultrametric 
```{r}
### get the distance from the root to each tip
tip.heights<-distRoot(tree)


### see which is the most common tip height and how much to adjust tips to all be the same
(heights.summary<-table(round(tip.heights,5)))

real.tree.height<-as.numeric(names(which.max(heights.summary)))
over.under<-tip.heights-real.tree.height


## extract all terminal edges for tips that do not have the final height we want:
tip.ids <- tree$edge[, 2] <= Ntip(tree)
terminal.edges <- tree$edge.length[tip.ids]


## add/subtract the extra length from the terminal branches
corrected.terminal.edges<-terminal.edges-over.under

## change the termnial edges in the phylo object
tree$edge.length[tip.ids]<-corrected.terminal.edges

```

#### Final test

```{r}
plot(tree, show.tip.label = F, main="its ultrametric!!")
is.ultrametric(tree)

```

#### Make it a function!

For convenience, we can now just throw it all together into a function and apply it to any tree we like.
```{r}
be.ultrametric<-function(phy) {
  
  tip.heights<-distRoot(phy)
  heights.summary<-table(round(tip.heights,5))
  real.tree.height<-as.numeric(names(which.max(heights.summary)))
  over.under<-tip.heights-real.tree.height
  tip.ids <- phy$edge[, 2] <= Ntip(phy)
  terminal.edges <- phy$edge.length[tip.ids]
  corrected.terminal.edges<-terminal.edges-over.under
  phy$edge.length[tip.ids]<-corrected.terminal.edges
  return(phy)
  
}
```


#### Worked example

I ran into this problem when I tried to use amphibian trees from the [Pyron 2014](https://academic.oup.com/sysbio/article/63/5/779/2847944) paper.

```{r}
amph.tree<-read.tree("https://datadryad.org/bitstream/handle/10255/dryad.63062/amph_shl_dates_frogs.tre?sequence=1")

# is it ultrametric?
is.ultrametric(amph.tree)

# strange... it looks pretty damn ultrametric
plot(amph.tree, show.tip.label = F)

# lets make it ultrametric then:
amph.tree.corrected<-be.ultrametric(phy=amph.tree)

# Boom! ultrametric!
plot(amph.tree.corrected, show.tip.label = F)
is.ultrametric(amph.tree.corrected)


```

#### Thoughts...

For large trees this code is pretty slow. Most of the heavy-lifting is done by *adephylo*'s `distRoot()` function. I wonder if there is a better way to do this similar to the inner workings of *ape*'s `is.ultrametric()` function. Drop me a line by starting a [github issue](https://github.com/hcliedtke/R-scrapheap/issues) related to this repository.

