---
title: "Trait evolution using mvMORPH"
author: "Christoph Liedtke"
date: "August, 2019"
output: 
  html_document:
    self_contained: true
    toc: True # table of content true
    toc_float:
      collapsed: false
      smooth_scroll: false
    depth: 4  # upto three depths of headings (specified by #, ##, ### etc.)
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Intro

Recently, we wanted to test whether a specific trait evolves under different evolutionary models for different subclades of a tree. In our case [genome size in amphibians](https://www.nature.com/articles/s41559-018-0674-4). To fit Brownian motion and Ornstein-Uhlenbeck models, we used Julien Clavel's [mvMORPH](https://github.com/JClavel/mvMORPH) R package.  

Tree: pruned amphibian tree originally from [Pyron 2014](https://academic.oup.com/sysbio/article/63/5/779/2847944)  
Data: genome sizes (c-values) for amphibian species from [Liedtke et al. 2018](https://www.nature.com/articles/s41559-018-0674-4)


# Loading Data and Libraries
```{r message=FALSE}
library(mvMORPH)
# install from CRAN or most recent release from github: https://github.com/JClavel/mvMORPH
# make sure dependencies are installed too!

```

```{r}
### load data:
amphibia_cval<-read.csv("./data/amphibia_cval.csv", row.names = 1, sep=";")

### load tree:
amphibia.tree<-read.tree("./data/amphibia.tre")

### match tree and data order
amphibia_cval<-amphibia_cval[amphibia.tree$tip.label,]
all(rownames(amphibia_cval)==amphibia.tree$tip.label)

### log transform variable to be normally distributed
amphibia_logC<-log10(amphibia_cval$median_c); names(amphibia_logC)<-rownames(amphibia_cval) # log10 transform data

### plot phenogram to vizualize trait distribution
phenogram(amphibia.tree, amphibia_logC, ftype = "off", ylab="genome size [log10(c-value)]")

```

# Setting up hypotheses

By looking at the phylogenetic distribution of the data, it seemed like salamanders were off doing their own thing with really large genomes (that isolated clade, all with large genomes) we wanted to test whether a single trait model for the whole tree is really best in this case, or whether fitting a multiple optimum/parameters model would be more just. This way we could also test the hypothesis of whether each of the three amphibian orders (Frogs, Salamanders and Caecilians) have evolved under different sets of parameters.  

To set up these hypotheses, we created a simmap object using (phytools)[http://blog.phytools.org/], a dependency of mvMORPH, 'painting' (i.e. annotating) subclades of the tree that we want to model different processes for.

```{r}
# make a colour scheme to use for vizualizing clades (in this case the three amphibian orders)
cols<-c("black","deepskyblue3","chartreuse3","gold"); names(cols)<-c(1,"caudata","anura","gymno")

# make a simmap object with two 'regimes', comparing salamanders to non-salamanders
two_regimes_tree<-paintSubTree(amphibia.tree,
                               node=getMRCA(amphibia.tree, rownames(amphibia_cval)[amphibia_cval$order=="Urodela"]),
                               state="caudata",
                               stem=T)
plotSimmap(two_regimes_tree,cols,lwd=2,pts=F, ftype="off")

# make a simmap object with three 'regimes' where each of the three amphibian orders is treated separately
three_regimes_tree<-paintSubTree(amphibia.tree,
                                 node=getMRCA(amphibia.tree, rownames(amphibia_cval)[amphibia_cval$order=="Gymnophiona"]),
                                 state="gymno",
                                 stem=T,
                                 anc.state = "gymno")
three_regimes_tree<-paintSubTree(three_regimes_tree,
                                 node=getMRCA(amphibia.tree,rownames(amphibia_cval)[amphibia_cval$order=="Urodela"]),
                                 state="caudata",
                                 stem=T)
three_regimes_tree<-paintSubTree(three_regimes_tree,
                                 node=getMRCA(amphibia.tree,rownames(amphibia_cval)[amphibia_cval$order=="Anura"]),
                                 state="anura",
                                 stem=T)


plotSimmap(three_regimes_tree,cols,lwd=2,pts=F, ftype="off")
```

# Fitting models

With the clades of interest defined, we can now fit BM and OU models. **note: as of late, mvMORPH also allows fitting early burst and shift models too!**

```{r}
# single optimum/parameter set:
fit_bm1<-mvBM(amphibia.tree, amphibia_logC,model="BM1",echo = F)
fit_ou1<-mvOU(amphibia.tree, amphibia_logC,model="OU1",echo = F)


# two optima/parameter sets:
fit_bm2<-mvBM(two_regimes_tree,amphibia_logC,model="BMM", param=list(smean=F),echo = F)
fit_ou2 <-mvOU(two_regimes_tree,amphibia_logC, model="OUM", param=list(root=F),echo = F)

# three optima/parameter sets:
fit_bm3<-mvBM(three_regimes_tree,amphibia_logC,model="BMM", param=list(smean=F),echo = T)
fit_ou3 <-mvOU(three_regimes_tree,amphibia_logC, model="OUM", param=list(root=F),echo = F)


```

## Compare model fit

Various things can be done here, for example we can calculate Akaike weights to compare models or we can perform likelihood ratio tests for each type of models (BM or OU)

```{r warnings=F}
## compare model fit using Akaike weights:
results<-list(fit_bm1,fit_ou1, fit_bm2, fit_ou2,fit_bm3,fit_ou3)
results<-aicw(results, aicc=TRUE)
results # note, the model names used can be checked in each of the model fit objects.

# Test significance with LRT
LRT(fit_bm1, fit_bm2)
LRT(fit_bm1, fit_bm3)
LRT(fit_bm2,fit_bm3) 

LRT(fit_ou1, fit_ou2)
LRT(fit_ou1, fit_ou3) 
LRT(fit_ou2,fit_ou3)
```

The AIC comparison suggests that the best model was a Brownian motion model with multiple sets of parameters, one for salamanders and one for everything else ('BMM default 3 model' in the Akeike weights table), but the likelihood ratio test was less clear about this, suggesting that that the 2 and 3 parameter sets performed equally well (or at least not significantly better or worse).  

# Ancestral state reconstruction

Seeing as the 2 and 3 parameter sets performed best, we are going to reconstruct ancestral states with these. Although BM outperformed OU, just as an example, I will reconstruct ancestral states with both.

```{r}
# BM (2 parameter model)
bm2.asr.fit<-mvBM(two_regimes_tree,amphibia_logC,model="BMM", param=list(smean=F)) 
bm2.asr.estim<-estim(two_regimes_tree, amphibia_logC, bm2.asr.fit, asr=TRUE)

# BM (3 parameter model)
bm3.asr.fit<-mvBM(three_regimes_tree,amphibia_logC,model="BMM", param=list(smean=F)) 
bm3.asr.estim<-estim(three_regimes_tree, amphibia_logC, bm3.asr.fit, asr=TRUE)


# OU (2 parameter model)
ou2.asr.fit<-mvOU(two_regimes_tree,amphibia_logC,model="OUM", param=list(root=F)) 
ou2.asr.estim<-estim(two_regimes_tree, amphibia_logC, ou2.asr.fit, asr=TRUE)

# OU (3 parameter model)
ou3.asr.fit<-mvOU(three_regimes_tree,amphibia_logC,model="OUM", param=list(root=F)) 
ou3.asr.estim<-estim(three_regimes_tree, amphibia_logC, ou3.asr.fit, asr=TRUE)


```

We can now visualize these reconstructions using phenograms for example:

```{r fig.height = 10, fig.width = 10}
## BM
#2 means
bm2.asr.states<-10^c(amphibia_logC,bm2.asr.estim$estimates) # The key here is to paste the tip states together with the node states estimated by mvMORPH as one vector. Here I am converting log10() traits back to real space so that I can plot a y axis on a log scale. this is not necesary, but I think it is more easily interpreted this way.
names(bm2.asr.states)[names(bm2.asr.states)==""]<-which(names(bm2.asr.states)=="") # some nodes don't have node numbers, which confuses phenogram() so this is just to make sure they are all numbered nicely

par(mfrow=c(2,2))
par(mar=c(4,4,1,1))
phenogram(tree=two_regimes_tree,x=bm2.asr.states, ftype="off", ylim=c(1,120), log="y", colors=cols)

#3 means
bm3.asr.states<-10^c(amphibia_logC,bm3.asr.estim$estimates)
names(bm3.asr.states)[names(bm3.asr.states)==""]<-which(names(bm3.asr.states)=="")

par(mar=c(4,4,1,1))
phenogram(tree=three_regimes_tree,x=bm3.asr.states, ftype="off", ylim=c(1,120), log="y", colors=cols)

## OU
#2 optima
ou2.asr.states<-10^c(amphibia_logC,ou2.asr.estim$estimates)
names(ou2.asr.states)[names(ou2.asr.states)==""]<-which(names(ou2.asr.states)=="")

par(mar=c(4,4,1,1))
phenogram(tree=two_regimes_tree,x=ou2.asr.states, fsize=0.0001, spread.labels=F, ylim=c(1,120), log="y", colors=cols, ftype="reg")

#3 optima
ou3.asr.states<-10^c(amphibia_logC,ou3.asr.estim$estimates)
names(ou3.asr.states)[names(ou3.asr.states)==""]<-which(names(ou3.asr.states)=="")

par(mar=c(4,4,1,1))
phenogram(tree=three_regimes_tree,x=ou3.asr.states, fsize=0.0001, spread.labels=F, ylim=c(1,120), log="y", colors=cols, ftype="reg")

```

# Simulating traits

Now that we have established the best model(s), we might also want to see whether any species or clades in particular may be outliers, i.e. showing trait histories diverging from the expected. To do this, we can simulate trait histories under the given models to get a distribution of trait values per species.

```{r fig.height = 15, fig.width = 10}
# Simulate
simul_bm2<-mvSIM(two_regimes_tree,nsim=10000, model="BMM",param=fit_bm2)

# Find species that fall outside 90% confidence intervals
simul_bm2.ci<-t(apply(FUN=quantile, c(0.05, 0.95), MARGIN = 1, X = simul_bm2))
simul_bm2.outliers<-which(simul_bm2.ci[,1] > amphibia_logC | simul_bm2.ci[,2] < amphibia_logC)
simul_bm2.outliers

# plot expected distributions and true trait values
par(mfrow=c(4,2))
for(i in 1:length(simul_bm2.outliers)){
  hist(simul_bm2[names(simul_bm2.outliers)[i],], main=names(simul_bm2.outliers)[i], cex=0.5, xlab="trait",las=1, border=NA, col="grey80")
  abline(v=quantile(simul_bm2[names(simul_bm2.outliers)[i],], c(0.05, 0.95)), col="blue", lwd=1, lty=2) # quantile bands
  abline(v=amphibia_logC[names(simul_bm2.outliers)[i]], col="red", lwd=2) # true value
}
```


## Simulating ancestral states

I couldn't find a way to get ancestral states for trait simulations (maybe there is a good reason for this), and so my work around has been to just perform ancestral state reconstructions on the simulated states. This seems somewhat circular to me, and so I want to put a big *disclaimer* here that maybe this isn't the best way to go about doing things.

```{r}
node.sim<-list()
## NOTE: here i am just using the first 100 simulations, but ideally you would run the loop for 1:ncol(simul_bm2)
for(i in 1:100){
  node.sim[[i]]<-estim(two_regimes_tree, simul_bm2[,i], bm2.asr.fit, asr=TRUE)$estimates[,1]
}
node.sim<-as.data.frame(t(do.call(rbind, node.sim)))

# Now you have a table of node states for each simulated trait set (simulations as columns)
# the root note will inevitably be the same for all simulations as this is the nature of a BM model:
node.sim[1,]

```


We could now look at the individual distributions per node if we are interested in a particular one, for example what would be the distribution of traits for the most recent common ancestor of salamanders:

```{r}
hist(as.numeric(node.sim["488",]), main="Simulated state for Salamander MRCA", xlab=NA, col="grey80", border=NA)
```

Using the `ggtree` package, we could also plot these histograms directly on the nodes of the tree. To do this, I borrowed a bit of code from [here](http://www.randigriffin.com/2017/05/11/primate-phylogeny-ggtree.html).
```{r fig.height = 12, fig.width = 10, warning = F}
library(ggtree)
library(ggplot2)

# make ggtee object
gtree<-ggtree(amphibia.tree)

# define vector with node numbers as character strings for which we want to plot the histograms
nodes <- c("466","488","657","694")

# make the node annotation object
pd <- as.data.frame(node.sim[nodes,])
pdplots <- apply(pd, 1, function(y) {
  ggplot(data.frame(y=y), aes(y, fill=..x..)) +
    geom_histogram(binwidth = 0.1, alpha=0.75) +
    xlim(range(unlist(pd))) +
    scale_fill_gradient(low = "blue", high = "red") +
    theme_inset() +
    theme(axis.line.x = element_line(color="black", size = 0.5),
          axis.text.x = element_text(size=8))
})

# plot tree and nodes
inset(gtree, pdplots, width=0.5, height=0.25, vjust=-1)
```



